---
title: "Data Visualization in R"
author: "Rose Mariyappan"
date: "April 21, 2019"
output: 
  html_document: 
    toc: true
    number_sections: true
---

# Base Graphics

This section of the document contains useful code snippets related to *R*'s default graphics system, base graphics. Reference includes:

* [DataCamp](https://www.datacamp.com/home)

## Exploratory & Explanatory Plots

Creating exploratory plot for all variables in a data frame - whiteside
```{r eval = FALSE}
plot(whiteside)
```
Explanatory Scatterplot - passing x, y axis variables and titles for sharing with others
```{r eval = FALSE}
plot(whiteside$Temp, whiteside$Gas, xlab="Outside temperature", ylab="Heating gas consumption")
title(main="heating gas consumption versus outside temperature")
```
Adding details to a plot using point shapes, color, and reference lines
Plot Max.Price vs. Price as red triangles
Add Min.Price vs. Price as blue circles
Add an equality reference line with abline()
```{r eval = FALSE}
plot(Cars93$Price, Cars93$Max.Price, pch=17, col="red")
points(Cars93$Price, Cars93$Min.Price, pch=16, col="blue")
abline(a = 0, b = 1, lty = 2)
```
Creating multiple plot arrays
Set up the side-by-side plot array
First plot: brain vs. body in its original form
Add the first title
Second plot: log-log plot of brain vs. body
Add the second title
```{r eval = FALSE}
par(mfrow = c(1,2))
plot(Animals2$body, Animals2$brain)
title("Original representation")
plot(Animals2$body, Animals2$brain, log="xy")
title("Log-log plot")
```
Create the pie chart
```{r eval = FALSE}
pie(tbl)
title("Pie chart")
```
Create the barplot with perpendicular, half-sized labels
```{r eval = FALSE}
barplot(tbl, las = 2, cex.names = 0.5)
title("Bar chart")
```
## Different Plot Types

Create a histogram of counts with hist()
```{r eval = FALSE}
hist(Cars93$Horsepower, main="hist() plot")
```
Create a normalized histogram with truehist() - from the MASS package and scales these counts to give an estimate of the probability density.
```{r eval = FALSE}
truehist(Cars93$Horsepower, main="truehist() plot")
```
Add the density curve to the histogram
```{r eval = FALSE}
lines(density(weights))
```
The quantile-quantile plot, or QQ-plot to know whether Gaussian distribution assumption is reasonable for data
Show the normal QQ-plot of the chick weights
```{r eval = FALSE}
qqPlot(weights)
```
sunflowerplot() function for repeated numerical data
```{r eval = FALSE}
sunflowerplot(Boston$rad ~ Boston$zn)
title("Sunflower plot")
```
The boxplot() function shows how the distribution of a numerical variable y differs across the unique levels of a second variable, x
Create a variable-width boxplot with log y-axis & horizontal labels
```{r eval = FALSE}
boxplot(Boston$crim ~ Boston$rad, varwidth = TRUE, log="y", las=1)
title("Crime rate vs. radial highway index")
```
A mosaic plot may be viewed as a scatterplot between categorical variables
```{r eval = FALSE}
mosaicplot(carb ~ cyl, data=mtcars)
```
correlation matrices with the corrplot() function
Load the corrplot library for the corrplot() function
Compute the correlation matrix for these variables
Generate the correlation ellipse plot
```{r eval = FALSE}
library(corrplot)
corrMat <- cor(numericalVars)
corrplot(corrMat, method="ellipse")
```
Build a decision tree model using the rpart() function 
Load the rpart library
Fit an rpart model to predict medv from all other Boston variables
Plot the structure of this decision tree model
Add labels to this plot
```{r eval = FALSE}
library(rpart)
tree_model <- rpart(medv ~., data = Boston)
plot(tree_model)
text(tree_model, cex=0.7)
```
## Plot Options

Add text giving names of cars next to data points - labels for x-y value pair
Add car names, offset from points, with larger bold text
Add "Before" text in blue, rotated 30 degrees, 80% size
Add "After" text in red, rotated -20 degrees, 80% size
```{r eval = FALSE}
text(x = Cars93$Horsepower[index3], 
     y = Cars93$MPG.city[index3],
     labels = Cars93$Make[index3], adj = 0)
text(Cars93$Horsepower[index3], Cars93$MPG.city[index3], label=Cars93$Make[index3], font=4, adj = -0.2, cex=1.2)
text(x = whiteside$Temp[indexB], y = whiteside$Gas[indexB],
     labels = "Before", col = "blue", srt = 30, cex = 0.8)
text(x=whiteside$Temp[indexA], y=whiteside$Gas[indexA], labels = "After", col="red", srt=-20, cex=0.8)
```
Add legend that identifies points as "Before" and "After"
```{r eval = FALSE}
legend("topright", pch = c(17, 1), 
       legend = c("Before", "After"))
```
Custom axes with axis()
Create a boxplot of sugars by shelf value, without axes
Add a default y-axis to the left of the boxplot
Add an x-axis below the plot, labelled 1, 2, and 3
Add a second x-axis above the plot
```{r eval = FALSE}
boxplot(UScereal$sugars~UScereal$shelf, axes=FALSE)
axis(side = 2)
axis(side = 1, at = c(1, 2, 3))
axis(side = 3, at = c(1, 2, 3), labels=c("floor", "middle", "top"))
```
Call supsmu() to generate a smooth trend curve, with bass=10
Add this trend curve to the plot
Add this trend curve as a heavy, dotted line
```{r eval = FALSE}
trend1 <- supsmu(Cars93$Horsepower, Cars93$MPG.city, bass=10)
lines(trend1)
lines(trend2, lty=3, lwd=2)
```
Common Scaling
Define common x and y limits for the four plots
Plot y1 vs. x1 with common x and y limits, labels & title. Similarly plot other datasets
```{r eval = FALSE}
xmin <- min(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4)
xmax <- max(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4)
ymin <- min(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
ymax <- max(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
plot(anscombe$x1, anscombe$y1,
     xlim = c(xmin, xmax),
     ylim = c(ymin, ymax),
     xlab = "x value", ylab = "y value",
     main = "First dataset")
```
## Plot Layouts

Constructing and displaying layout matrices
Use the matrix function to create a matrix with three rows and two columns
Call the layout() function to set up the plot array
Show where the three plots will go 
Create 3 plots to view
```{r eval = FALSE}
layoutMatrix <- matrix(
  c(
    0, 1,
    2, 0,
    0, 3
  ), 
  byrow = TRUE, 
  nrow = 3
)
layout(layoutMatrix)
layout.show(n=3)
plot(whiteside$Temp[indexB], whiteside$Gas[indexB], ylim = c(0, 8))
title("Before data only")
plot(whiteside$Temp, whiteside$Gas, ylim = c(0, 8))
title("Complete dataset")
plot(whiteside$Temp[indexA], whiteside$Gas[indexA], ylim = c(0, 8))
title("After data only")
```
Creating arrays with different sized plots
Create row1, row2, and layoutVector
Convert layoutVector into layoutMatrix
Set up the plot array
Plot scatterplot
Plot sunflower plot
```{r eval = FALSE}
row1 <- c(1,0,0)
row2 <- c(0,2,2)
layoutVector <- c(row1,row2,row2)
layoutMatrix <- matrix(layoutVector, byrow = TRUE, nrow = 3)
layout(layoutMatrix)
plot(Boston$rad, Boston$zn)
sunflowerplot(Boston$rad, Boston$zn)
```
Generate a horizontal barplot of these frequencies
Add names labels with text()
Add count labels with text()
```{r eval = FALSE}
barplot(tbl, horiz =TRUE,
                col = "transparent",
                names.arg = "")
text(20, mids, names(tbl))
text(35, mids, as.numeric(tbl))
```
Call symbols() to create the default bubbleplot
The symbols() function allows us to extend scatterplots to show the influence of other variables.
```{r eval = FALSE}
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price),
        inches = 0.1)
```
## Saving Plot Files

Saving plot results as files
Call png() with the name of the file we want to create
Re-create the plot from the last exercise
Save our file and return to our interactive session
Verify that we have created the file
```{r eval = FALSE}
png("bubbleplot.png")
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price),
        inches = 0.1)
dev.off()
list.files(pattern = "png")
```
## Plot colors

Using color to enhance a bubbleplot
Iliinsky and Steele color name vector
Create the `cylinderLevels` variable
Create the colored bubbleplot
Using color to enhance stacked barplots
```{r eval = FALSE}
IScolors <- c("red", "green", "yellow", "blue",
              "black", "white", "pink", "cyan",
              "gray", "orange", "brown", "purple")
cylinderLevels <- as.numeric(Cars93$Cylinders)
symbols(Cars93$Horsepower, Cars93$MPG.city, 
        circles = cylinderLevels, inches = 0.2, 
        bg = IScolors[cylinderLevels])
tbl <- table(Cars93$Cylinders, Cars93$Origin)
barplot(tbl, col=IScolors)
```

# Grammer of graphics - ggplot2

This section of the document contains useful code snippets related to ggplot2 in *R* programming language. Reference includes:

* [DataCamp](https://www.datacamp.com/home)

## Exploring ggplot2

Load the ggplot2 package and scatter plot, here cyl is treated as factor
```{r eval = FALSE}
library(ggplot2)
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_point()
```
Add geom_point() and geom_smooth() with + to draw smooth line
```{r eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point() +
geom_smooth()
```
Add color for clarity. Plot points with argument alpha to make the points 60% transparent/40% visible..
```{r eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.4)
```
Include the lines of the linear models, per cyl & lm for the entire dataset in its whole
```{r eval = FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
  geom_point() + # Copy from Plot 2
  geom_smooth(method = "lm", se = FALSE) + # Copy from Plot 2
  geom_smooth(method = "lm", se = FALSE, lty = 2, aes(group = 1)) 
```
column-wise comparison of measure values - length & width specified in facet_grid, colored by Part values
```{r eval = FALSE}
ggplot(iris.tidy, aes(x = Species, y = Value, col = Part)) +
  geom_jitter() +
  facet_grid(. ~ Measure)
```
## Aesthetics

Map wt to x, mpg to y and cyl to col & change shape and size of the points
```{r eval = FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(shape = 1, size = 4)
```
cyl to fill(for inside of point) & am to color for outside 
```{r eval = FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl, col = am)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)
```
Map cyl to label
```{r eval = FALSE}
ggplot(mtcars, aes(x=wt, y=mpg, label = cyl), main="mpg described by wt") +
geom_text()
```
Set the fill aesthetic; color, size and shape attributes
```{r eval = FALSE}
ggplot(mtcars, aes(x=wt, y=mpg, fill=cyl)) +
geom_point(color = my_color, size=10, shape = 23)
```
Expand to draw text with label rownames(mtcars) and color red
```{r eval = FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, fill = cyl)) +
geom_text(label = rownames(mtcars), color = "red" )
```
Use the "stack", "fill" and "dodge" positions
Add geom (position = "stack" by default)
Fill - show proportion
Dodging - principles of similarity and proximity
Clean up the axes with scale_ functions
```{r eval = FALSE}
cyl.am + 
  geom_bar()
cyl.am + 
  geom_bar(position = "fill") 
cyl.am +
  geom_bar(position = "dodge") 
val = c("#E41A1C", "#377EB8")
lab = c("Manual", "Automatic")
cyl.am +
  geom_bar(position = "dodge") +
  scale_x_discrete("Cylinders") + 
  scale_y_continuous("Number") +
  scale_fill_manual("Transmission", 
                    values = val,
                    labels = lab) 
```
Add function to change y axis limits
```{r eval = FALSE}
ggplot(mtcars, aes(x = mpg, y = 0)) +
  geom_jitter() +
  scale_y_continuous(limits = c(-2, 2))
```
Overplotting - Point shape and transparency
```{r eval = FALSE}
ggplot(mtcars, aes(x=wt, y=mpg, col = cyl)) + 
geom_point(size = 4, shape = 1)
ggplot(mtcars, aes(x=wt, y=mpg, col = cyl)) + 
geom_point(size = 4, alpha = 0.6)
```
Overplotting - Use position = "jitter" if all individual values line up on a single axis within each clarity category
```{r eval = FALSE}
ggplot(diamonds, aes(x=clarity, y=carat, color=price)) +
geom_point(alpha = 0.5, position="jitter")
```
## Different plot types

Scatter plot and jittering
```{r eval = FALSE}
ggplot(Vocab, aes(x=education, y=vocabulary)) +
geom_jitter(alpha = 0.2, shape = 1)
```
Histograms - fill to custom color, adjust binwidth, plot by density
```{r eval = FALSE}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill="#377EB8")
```
Barplot - Use posn_d as position and adjust alpha to 0.6
```{r eval = FALSE}
posn_d <- position_dodge(width = 0.2)
ggplot(mtcars, aes(x = cyl, fill = am)) +
  geom_bar(position = posn_d, alpha=0.6)
```
Barplot - Plot education on x and vocabulary on fill, use the default brewed color palette
```{r eval = FALSE}
ggplot(Vocab, aes(x = education, fill = vocabulary)) +
  geom_bar(position="fill") +
  scale_fill_brewer()
```
Lineplots - plot to represent the fraction of total population that is unemployed
```{r eval = FALSE}
ggplot(economics, aes(x = date, y = unemploy/pop)) +
geom_line()
```
Lineplots - Expand economics to draw the recess periods using geom_rect()
```{r eval = FALSE}
ggplot(economics, aes(x = date, y = unemploy/pop)) +
  geom_rect(data = recess,
         aes(xmin = begin, xmax = end, ymin = -Inf, ymax = +Inf),
         inherit.aes = FALSE, fill = "red", alpha = 0.2) +
  geom_line()
```
Multiple time series
```{r eval = FALSE}
ggplot(fish.tidy, aes(x = Year, y = Capture, color=Species)) +
geom_line()
```
qplot for quick & dirty exploratory plots
```{r eval = FALSE}
qplot(wt, mpg, data = mtcars)
qplot(wt, mpg, data = mtcars, col = qsec)
qplot(factor(cyl), factor(vs), data=mtcars, geom="jitter")
```

# Interactive graphics

This section of the document contains useful code snippets related to interactive graphics in *R* programming language. Reference include:

* [DataCamp](https://www.datacamp.com/home)

## Shiny for interactive reports
Shiny is an R package that uses R to build interactive web apps such as data explorers and dashboards. You can add shiny components to an R Markdown file to make an interactive document.

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r eval = FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),

  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")

  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Interactive ggvis graphics
ggvis provides a number of ways to enhance plots with interactivity. ggvis relies on the shiny framework to create interactivity.

The density plot below allows users to set the kernel and bandwidth of the plot.
```{r eval = FALSE}
library(ggvis)

mtcars %>% ggvis(x = ~wt) %>%
    layer_densities(
      adjust = input_slider(.1, 2, value = 1, step = .1, label = "Bandwidth adjustment"),
      kernel = input_select(
        c("Gaussian" = "gaussian",
          "Epanechnikov" = "epanechnikov",
          "Rectangular" = "rectangular",
          "Triangular" = "triangular",
          "Biweight" = "biweight",
          "Cosine" = "cosine",
          "Optcosine" = "optcosine"),
        label = "Kernel")
    )
```